---
title: "netinf-quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{netinf-quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

# Quick Start

`netinf` identifies **influential cases** in psychological network analysis.
It reuses the same bootstrap samples of `bootnet` for influence
diagnostics.

For the theory and formulas behind each step, see the
[Detailed Guide](netinf-detailedguide.html).

## Setup

```{r}
library(netinf)
```

## The data

The package ships with a simulated dataset `test_data`: 100 cases × 9
variables. The last case (row 100) is an artificially constructed
influential case.

```{r}
data("test_data", package = "netinf")
```

## The all-in-one function

For most users, `influence_diagnostic()` runs the whole workflow in one call:

```{r, eval = FALSE}
diag <- influence_diagnostic(
  data               = test_data,
  nBoots             = 5000,
  centrality_metrics = c("Strength", "Closeness", "Betweenness"), # Select the statistics to calculate the generalized Cook distance.
  default            = "EBICglasso",
  verbose            = FALSE
)

```

The output contains three parts:

- **Phase 1**: empirical influence values per case
- **Phase 2**: leave-one-out validation of top candidates
- **Phase 3**: centrality gCD (generalized Cook's distance)

## Manual workflow

If you want to run each step yourself, the four functions form a pipeline.

### Step 1: bootstrap with case indices

`bootnet_with_indices()` runs `bootnet::bootnet()` and additionally stores
the case index drawn in each bootstrap sample.

```{r}
set.seed(1)
boot_res <- bootnet_with_indices(
  test_data,
  nBoots  = 5000,
  default = "EBICglasso"
)

```

### Step 2: empirical influence per case

`calculate_empirical_influence()` turns the bootstrap indices into a numeric
influence value per case.

```{r}
empic <- calculate_empirical_influence(boot_res)

print(empic, n = 10)   # 10 cases with largest absolute value
```

### Step 3: leave-one-out validation

`leave_one_out_analysis()` removes the selected candidates and re-estimates the
network to see what changes.

There are three ways to select cases:

- `top_n`: top N by absolute influence from `calculate_empirical_influence()`
- `threshold`: cases above a cutoff (with `direction` to pick the tail) from `calculate_empirical_influence()`
- `remove_cases`: manually specified IDs

**Option 1 — top N:**

```{r}
loo_res <- leave_one_out_analysis(
  influence_result = empic,
  data             = test_data,
  top_n            = 3,
  default          = "EBICglasso"
)

loo_res
```

**Option 2 — by threshold:**

```{r}
loo_thr <- leave_one_out_analysis(
  influence_result = empic,
  data             = test_data,
  threshold        = 8,
  direction        = "both",        # "both" | "positive" | "negative"
  default          = "EBICglasso"
)

loo_thr
```

**Option 3 — manually remove specific cases (multiple → leave-multiple-out):**

```{r}
lmo_res <- leave_one_out_analysis(
  influence_result = empic,
  data             = test_data,
  remove_cases     = c(3, 7, 12),
  default          = "EBICglasso"
)

lmo_res
```

### Step 4: centrality gCD

`calculate_centrality_gCD()` quantifies how much each case shifts the
centrality vector.

```{r}
gcd_res <- calculate_centrality_gCD(
  data        = test_data,
  boot_result = boot_res,
  case_ids    = c(100, 75, 50, 25),
  metric      = c("Strength", "Closeness", "Betweenness"),
  default     = "EBICglasso"
)

gcd_res
```

## Visualization

Three plot functions work with either the all-in-one object or individual
step outputs.

### influence_plot

```{r}
influence_plot(empic, largest_x = 3)
```

### gcds_plot

```{r}
gcds_plot(gcd_res, metrics = "Strength")
```

### gcds_md_plot

```{r}
gcds_md_plot(gcd_res, data = test_data, metric = "Strength")
```

## Tips

- If you already ran `bootnet_with_indices()`, pass the result to
  `influence_diagnostic()` to skip Step 1:
- Only `default = "EBICglasso"` is currently supported.
- `calculate_centrality_gCD()` needs `case_ids`; these can be any cases you
  want to diagnose—they do not have to come from `leave_one_out_analysis()`.

## Next steps

- [Detailed Guide](netinf-detailedguide.html) — theory, formulas, and
  interpretation for each step.
- `?influence_diagnostic` — function reference.
